A Plasma Torus Around a Young Low-Mass Star

### 2025-04-23 paper/ms.tex
update

   while the spectrum lacks lithium (see \S~\ref{sec:stparams}).

@@

4: Keck/HIRES (see \S~\ref{subsec:halpha}).

@@

\subsection{Corotating Plasma Clumps}

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they demonstrate that corotating circumstellar plasma
clumps exist around at least one such star.  While starspots do
contribute smooth signals to the photometric variability, a
``starspot-only'' scenario to explain the complexity of the CPVs
\citep{Koen2021} is incompatible with our observations.  Scenarios in
which the circumstellar material is made only of dust are similarly
ruled out.  While starspots and dust may both be present, our HIRES
spectra require plasma with a significant population of hydrogen atoms
in the $n$=3 excited state, with no evidence for significant
higher-order excitations (see Appendix~\ref{subsec:specvar}).  Most of
this H$\alpha$ emission originates in ``clumps'' with size similar to
that of the star; radial ``spokes'' or azimuthally-extended ``arcs''
for the emitting material are ruled out by the observed H$\alpha$
velocity dispersion (see Appendix~\ref{subsec:halpha}).  The observed
H$\alpha$ luminosity suggests characteristic densities and masses for
the gaseous component of these clumps of $n_{\rm H} \sim
10^{11}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g (see
Appendix~\ref{subsec:gas}).  Dust, if present, is independently
constrained to have a total mass $10^{15}\,{\rm g} < M_{\rm dust} <
10^{17}\,{\rm g}$ (see Appendix~\ref{subsec:dust}).

%TODO: improve heading?
\subsection{Plasma Tori Could Explain Complex Periodic Variables}

While the H$\alpha$ emission appears consistent with clumps, the
photometric width of the complex flux dips in TIC\,141146667
specifically, and CPVs more broadly \citep[Figure~4 of][]{Bouma2024},
raise the possibility that azimuthally-distributed material may be the
norm for these objects.  In other words, a torus may be a more
accurate picture than a clump.

\citet{Townsend2005} outline the physics of how such structures can
exist: for stars with magnetospheric radii $R_{\rm m}$ that exceed
their corotation radii $R_{\rm c}$, the effective potential
experienced by a charged particle can have a local minimum outside
$R_{\rm c}$, enabling material to accumulate in a
centrifugally-supported magnetosphere
\citep{Petit2013,Daley-Yates2024}.  Generally speaking, such material
need neither transit nor be opaque in broadband optical light.  In
fact, Figure~\ref{fig:spec} suggests that the transits of the
H$\alpha$ emitting material cannot explain the entirety of the complex
photometric modulation;  the spectroscopic transits happen too fast,
and at orbital phases that only partially overlap with the photometric
modulations.

Nonetheless, \citet{Townsend2008} synthesized light curves for plasma
trapped in rigidly rotating magnetospheres of rapidly rotating stars;
and found that assuming a dipole magnetic geometry, one can obtain
W-shaped eclipses similar to those seen for TIC\,141146667.  See in
particular the
\href{http://user.astro.wisc.edu/~townsend/static.php?ref=rrm-movies#Download_Bundles}{movies}
associated with their work; for a $\omega/\omega_{\rm c}$$\approx$0.5,
edge-on system like TIC\,141146667, the eclipses have the correct
approximate shape, and the H$\alpha$ emission can similarly exhibit
double-peaked behavior.

>>>>

   while the spectrum lacks lithium (Appendix~\ref{sec:stparams}).

@@

4: Keck/HIRES (Appendix~\ref{subsec:halpha}).

@@

\subsection{Physical Properties of the Corotating Plasma Clumps}

% this paragraph: physical properties that can be inferred with
% confidence

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they demonstrate that corotating circumstellar plasma
clumps exist around at least one such star.  More specifically, our
HIRES spectra require plasma with a significant population of hydrogen
in the $n$=3 excited state, with minimal evidence for 
higher-order excitations (see Appendix~\ref{subsec:specvar}).  Most of
this H$\alpha$ emission originates in ``clumps'' with size similar to
that of the star; radial ``spokes'' or azimuthally-extended ``arcs''
for the {\it emitting material} are ruled out by the observed H$\alpha$
velocity dispersion (see Appendix~\ref{subsec:halpha}).  The H$\alpha$
line luminosity suggests characteristic densities and masses for the
gaseous component of these clumps of $n_{\rm H} \sim
10^{11}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g (see
Appendix~\ref{subsec:gas}).  Dust, if present, is independently
constrained to have a total mass $10^{15}\,{\rm g} < M_{\rm dust} <
10^{17}\,{\rm g}$ (see Appendix~\ref{subsec:dust}).

\subsection{Theoretical Context}

While the H$\alpha$ emission appears to mostly originate from clumps,
the the complex flux dip in TIC\,141146667 spans roughly half of each
cyle; in CPVs more broadly, dips typically last 5-50\% of each cycle
\citep{Bouma2024}.  This raises the possibility that
azimuthally-distributed material may be the norm for these objects.
In other words, a warped, azimuthally-asymmetric torus may be a more
accurate picture than a clump.

\citet{Townsend2005} outline the physics of how such structures might
exist for the azimuthally symmetric case.
A charged particle that orbits a 
star with a magnetospheric radius $R_{\rm m}$ that exceeds
its Keplerian corotation radius $R_{\rm c}$ experiences an effective potential
that has a local minimum outside
$R_{\rm c}$, which enables material to accumulate in these
centrifugally-supported magnetosphere
\citep{Petit2013,Daley-Yates2024}.  Generally speaking, such material
need neither transit nor be opaque in broadband optical light.  In
fact, Figure~\ref{fig:spec} suggests that the transits of the
H$\alpha$ emitting material cannot explain the entirety of the complex
photometric modulation;  the spectroscopic transits happen too fast,
and at orbital phases that only partially overlap with the photometric
modulations.

In a follow-up study, \citet{Townsend2008} synthesized light curves
for plasma trapped in rigidly rotating magnetospheres of rapidly
rotating stars; and found that assuming a dipole magnetic geometry,
one can obtain W-shaped eclipses similar to those seen for
TIC\,141146667.  See in particular the
\href{http://user.astro.wisc.edu/~townsend/static.php?ref=rrm-movies#Download_Bundles}{movies}
associated with their work; for a $\omega/\omega_{\rm c}$$\approx$0.5,
edge-on system like TIC\,141146667, the eclipses have the correct
approximate shape, and the H$\alpha$ emission can similarly exhibit
double-peaked behavior.

\subsection{Plasma Tori and Complex Photometric Morphology}

Microphysically, it is not obvious whether hydrogen alone can produce
the broadband opacities and chromatic flux variations seen in CPVs.
For detailed discussion of this point, see Section 5.6 in
\citet{Bouma2024}.
With that said, Figures~\ref{fig:fulllc} and~\ref{fig:spec} show that
at least {\it some} of the complex photometric dip for TIC\,141146667
is temporally coincident with the transit of the inner clump of
H$\alpha$ emitting material.
This could be because

However, the transit lasts $\approx$22\% of the

$\phi$=-0.1 to $\phi$=+0.12 (see
Appendix~\ref{subsec:halpha}).  This spectroscopic transit coincides with
the latter half of the complex eclipse feature in the TESS data.

 suggests that the transits of the
 cannot explain the entirety of the complex
photometric modulation;  the spectroscopic transits happen too fast,
and at orbital phases that only partially overlap with the photometric
modulations.

is not optically thick.
 discussed the general issue: CPV dips are chromatic,

In
fact, Figure~\ref{fig:spec} suggests that the transits of the
H$\alpha$ emitting material cannot explain the entirety of the complex
photometric modulation;  the spectroscopic transits happen too fast,
and at orbital phases that only partially overlap with the photometric
modulations.

While starspots do
contribute smooth signals to the photometric variability, a
``starspot-only'' scenario to explain the complexity of the CPVs
\citep{Koen2021} is incompatible with our observations.  Scenarios in
which the circumstellar material is made only of dust are similarly
ruled out.  While starspots and dust may both be present,
